---
title: "01: Summary Statistics"
author: "Herman Yu"
output:
  rmdformats::downcute
---

# Setup
```{r, warning=FALSE, message=FALSE}
options(scipen=999)
DATA_DIR <- Sys.getenv("DATA_DIR")
DATA_PATH <- paste0(DATA_DIR, "/store-sales-time-series/train.csv")

library(fpp3)
library(readr)
library(DT)
library(scales)
```

```{r, warning=FALSE, message=FALSE}
# load data
df <- read_csv(DATA_PATH)

df %>% 
  head(10) %>% 
  datatable()
```


# Data Integrity

## Dimensions

```{r}
df %>% 
  summarise(
    num_rows = n(),
    start_date = min(date),
    end_date = max(date),
    distinct_dates = n_distinct(date),
    distinct_families = n_distinct(family),
    distinct_stores = n_distinct(store_nbr)
  ) %>% 
  datatable()
```

The number of rows (3,000,888) exactly equals the product of 1,684 x 33 x 54; there is 1 row per unique combination of `date` x `family` x `store_nbr`.

## Missing Values

```{r}
df %>% 
  summarise(
    na_dates = sum(is.na(date)),
    na_family = sum(is.na(family)),
    na_stores = sum(is.na(store_nbr))
  )
```

## Gaps in Date Range

```{r}
tibble(
  date = seq(ymd("2013-01-01"), ymd("2017-08-15"), by = "day")
) %>% 
  left_join(
    df %>% 
      select(
        date
      ) %>% 
      mutate(
        is_in_data = 1
      )
    ,
    by = "date"
  ) %>% 
  filter(
    is.na(is_in_data)
  )
```

Christmas Day is not present in the data for all years.

# Product Family

```{r}
family_list <- df %>% 
  distinct(family) %>% 
  pull()

length(family_list)
```

## Time Series By Family

```{r, fig.height = 15, fig.width = 10}
df %>% 
  group_by(date, family) %>% 
  summarise(
    sales = sum(sales),
    .groups = "drop"
  ) %>% 
  ggplot(aes(date, sales)) + 
  geom_line() + 
  facet_wrap(~family, ncol = 3, scales = "free_y")
```

Notes:

* Many product family always have 0 or near 0 sales on New Years.
* Some product families only had sales starting from a certain point in time.
* Frozen foods, groceries, home and kitchen, liquor/beer/wine show consistent spikes during the end of the year.
* School supplies always spike in the Fall around August.

## Summary Of Family Sales

```{r}
df %>% 
  group_by(
    family
  ) %>% 
  summarise(
    total_sales = sum(sales),
    avg_daily_sales = mean(sales),
    days_on_promo = sum(onpromotion),
    avg_days_on_promo = mean(onpromotion)
  ) %>% 
  arrange(
    desc(total_sales)
  ) %>% 
  datatable()
```

```{r}
df %>% 
  filter(sales > 0) %>% 
  group_by(family) %>% 
  summarise(
    begin_date = min(date)
  ) %>% 
  arrange(
    desc(begin_date)
  ) %>% 
  datatable()
```

* Books sales volume. Books sales only start in October 2016.
* 10 products only start sales in early 2014.
* Remaining products all have sales in January 2013.

## Sales Mix By Family

```{r, fig.width = 12, fig.height = 5}
df %>% 
  group_by(date, family) %>% 
  summarise(
    sales = sum(sales),
    .groups = "drop"
  ) %>% 
  group_by(date) %>% 
  mutate(
    sales_percent = sales/sum(sales)
  ) %>% 
  ungroup() %>% 
  ggplot(aes(date, sales_percent, fill = as.factor(family))) + 
  geom_area()
```

```{r}
df %>% 
  group_by(date, family) %>% 
  summarise(
    sales = sum(sales),
    .groups = "drop"
  ) %>% 
  group_by(date) %>% 
  mutate(
    sales_percent = sales/sum(sales)
  ) %>% 
  ungroup() %>% 
  mutate(
    mean_sales_percent = mean(sales_percent)
  ) %>% 
  group_by(family) %>% 
  summarise(
    family_mix_stdev = sd(sales_percent),
    max_dev = max(abs(sales_percent - mean_sales_percent)),
    daily_sales = mean(sales)
  ) %>% 
  arrange(desc(family_mix_stdev)) %>% 
  datatable()
```


# Stores

## Time Series By Store

```{r, fig.height = 20, fig.width = 10}
df %>% 
  group_by(date, store_nbr) %>% 
  summarise(
    sales = sum(sales),
    .groups = "drop"
  ) %>% 
  ggplot(aes(date, sales)) + 
  geom_line() + 
  facet_wrap(~store_nbr, ncol = 3, scales = "free_y")
```

## Store Sales Summary

```{r}
df %>% 
  group_by(store_nbr) %>% 
  summarise(
    total_sales = sum(sales),
    avg_daily_sales = mean(sales),
    days_on_promo = sum(onpromotion),
    avg_days_on_promo = mean(onpromotion)
  ) %>% 
  arrange(
    desc(total_sales)
  ) %>% 
  datatable()
```

```{r}
df %>% 
  filter(sales > 0) %>% 
  group_by(store_nbr) %>% 
  summarise(
    begin_date = min(date)
  ) %>% 
  arrange(
    desc(begin_date)
  ) %>% 
  datatable()
```

## Sales Mix By Store

```{r, fig.width = 12, fig.height = 6}
df %>% 
  group_by(date, store_nbr) %>% 
  summarise(
    sales = sum(sales),
    .groups = "drop"
  ) %>% 
  group_by(date) %>% 
  mutate(
    sales_percent = sales/sum(sales)
  ) %>% 
  ungroup() %>% 
  ggplot(aes(date, sales_percent, fill = as.factor(store_nbr))) + 
  geom_area()
```


```{r}
df %>% 
  group_by(date, store_nbr) %>% 
  summarise(
    sales = sum(sales),
    .groups = "drop"
  ) %>% 
  group_by(date) %>% 
  mutate(
    sales_percent = sales/sum(sales)
  ) %>% 
  ungroup() %>% 
  mutate(
    mean_sales_percent = mean(sales_percent)
  ) %>% 
  group_by(store_nbr) %>% 
  summarise(
    store_mix_stdev = sd(sales_percent),
    min_dev = max(abs(sales_percent - mean_sales_percent)),
    daily_sales = mean(sales)
  ) %>% 
  arrange(desc(store_mix_stdev)) %>% 
  datatable()
```

---